Supp mat: time series paper

### 2017-08-15 writing/manuscript_draft.Rmd
major revisions to introduction

Title: Statistical power and detecting population trends: the need for long-term monitoring programs
\vspace{7 mm}

@@

To be submitted to: \emph{Biological Conservation} (Short Communication)
\vspace{3 mm}

@@

# Abstract
Long-term census data are necessary to better understand ecological processes and make management decisions. However, census data are often expensive, requiring a lot of time and resource. Little work has addressed the question, when is a time series of census data long enough to address a question of interest? Here, we explore two approaches to address this question: simulations and an empirical approach. We specifically determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends. Most studies, especially those less than 10-15 years, are probably under-powered and potentially misleading.

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017]. Long-term data are necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "long-term", most are still short. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017]. There are, of course, notable exceptions where long-term monitoring has occurred, for example the continuous plankton recorder [@Giron-Nava2017].

Naturally, one might ask, how long of a time series is actually necessary? This question has important implications for both research and management. Scientifically, we need to know the time series length required to address a specific question. Too short of a time series may lead to wrong conclusions given large natural year-to-year variability (citations). From a management perspective, it is important to understand when population trend over time is actually meaningful or not. The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary [@Gerber1999].

In determining the number of samples required for a particular experiment four quantities are intricately linked: statistical power (1-$\beta$), effect size, $\alpha$, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$), or false negative. In other words, it is a measure of how well you could detect a trend given it is actually significant. Prior to an experiment, one could set appropriate levels of power, significance threshold ($\alpha$), and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power [@Bolker2008]. Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determining sample size, simulation approaches can be used [@Gerrodette1987; @Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question [@Bolker2008]. In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions. In the main manuscript, we estimate the minimum time series length required to assess long-term changes in abundance via simple linear regression. We then examine 878 population time series to estimate this minimum time series length required for detecting population trends. We expand on the results in the supplmentary material by examining other questions of interest.

@@

counts <- table(pop_info$class,pop_info$min_time_for_power)
barplot(counts, main=" ",
        xlab=" ", ylab="", col=1:4,
        legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
        las=1,cex.names =1.2,cex.axis = 1.2,density=30,
        angle=c(20,40,60,80))

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=5}

# Plot minimum time required against trend strength, population variance, and autocorrelation  
  par(mfrow=c(3,1),mar=c(4,2,2,1),oma=c(0,4.5,0,0.5))
  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  mtext('| trend strength |',1,line=3,cex=1.2)
  mtext('(a)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
    mtext('population variance',1,line=3,cex=1.3)
    mtext('(b)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
      mtext('autocorrelation',1,line=3,cex=1.2)
       mtext('(c)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)

>>>>

Title: Minimum time required to detect population trends: the need for long-term monitoring programs
\vspace{7 mm}

@@

To be submitted to: \emph{Biological Conservation}
\vspace{3 mm}

@@

# Abstract
Long-term census data are necessary to better understand ecological processes and make management decisions. However, census data are often expensive, requiring a lot of time and resources. Little work has addressed the question: when is a time series of census data long enough to address a question of interest? <!--Here, we explore two approaches to address this question: simulations and an empirical approach.--> We determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.

@@

Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017]. Long-term data are necessary to both understand population dynamics and to assess species extinction risk (citations). Even though many time series may now be considered "long-term" (e.g. continuous plankton recorder @Giron-Nava2017), most are still short. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017].

How long of a time series is actually necessary? This question has important implications for both research and management. Scientists need to know the time series length required to address a specific question. A short time series may lead to wrong conclusions given large natural year-to-year variability (citations). Managers need to know when action is needed for a population. Therefore, they must understand when population trend over time is actually meaningful. The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary [@Gerber1999].

An important step in experimental design is to determine the number of samples required. For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. A type I error is a false positive, or incorrect rejection of a true null hypothesis. For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive. The false positive rate, or significance level ($\alpha$) is often set at 0.05. A type II error ($\beta$) is failure to detect a true trend, or failure to reject a false null hypothesis. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$). The effect size is a measure of the difference between two groups. Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.

For time series data, two general approaches to estimating sample size are appropriate. <!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g. t-test). However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015]. As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987]. Statistical power can then be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. For example, given a time series, when is the slope from linear regression significantly different from zero? In other words, when is the time series significantly increasing or decreasing? It is then possible to determine how power changes with a variable of interest. For example, time series are simulated for different lengths of time. From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].

In addition to using simulations, empirical time series can also be used. Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. In addition, @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

We use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions. We estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression. First, we estimate $T_{min}$ using a simulation approach. Then we examine 878 population time series to estimate $T_{min}$. In the supplementary material, we determine $T_{min}$ for related ideas: calculating long-term growth rates, using more complicated population models, varying statistical level and power, and the use of generalized additive models.

@@

counts <- table(pop_info$class,pop_info$min_time_for_power)
barplot(counts, main=" ",
        xlab=" ", ylab="", col=1:4,
        legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
        las=1,cex.names =1.2,cex.axis = 1.2,density=35,
        angle=c(20,40,60,80))

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}

# Plot minimum time required against trend strength, population variance, and autocorrelation  
  par(mfrow=c(3,1),mar=c(3,2,2,1),oma=c(1,4.5,0,0.5))
  plot(abs(pop_info$overall_trend),pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
  mtext('| trend strength |',1,line=3,cex=1.2)
  mtext('(a)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
    plot(pop_info$variance,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
    mtext('population variance',1,line=3,cex=1.2)
    mtext('(b)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
      plot(pop_info$autocorrelation,pop_info$min_time_for_power,pch=16,col=rgb(0.5,0.5,0.5,0.5),las=1,ylab='',xlab='',cex.axis=1.2)
      mtext('autocorrelation',1,line=3,cex=1.2)
       mtext('(c)',3,line=-1.5,adj = 0.985,cex=1.2,font = 2)
       mtext('minimum time required',2,line=2.5,cex=1.2,outer = T)

### 2017-08-15 writing/manuscript_draft.tex
major revisions to introduction

To be submitted to: \vspace{3 mm}

@@

Long-term census data are necessary to better understand ecological
processes and make management decisions. However, census data are often
expensive, requiring a lot of time and resource. Little work has
addressed the question, when is a time series of census data long enough
to address a question of interest? Here, we explore two approaches to
address this question: simulations and an empirical approach. We
specifically determine the minimum time series length required to
estimate significant increases or decreases in population abundance.
Importantly, we examine the ability to detect a trend with a set level
of statistical power, which is often neglected in ecological time series
analyses. Using a simulation approach, we demonstrate how the minimum
time series length required increases with weaker trends in abundance
and with higher population variability. In addition, we examine 878
populations of vertebrate species to determine the minimum time required
to detect changes in their abundance. We show that 10-15 years of
continuous monitoring are required in order to achieve a high level of
statistical power. Similar to simulation-approaches, the minimum time
required for field census data strongly depends on trend strength,
population variability, and temporal autocorrelation. These results
point to the importance of sampling populations over long periods of
time. We stress the need for verifying a study has enough statistical
power to accurately determine long-term population trends. Most studies,
especially those less than 10-15 years, are probably under-powered and
potentially misleading.

  \renewcommand{\thesection}{\arabic{section}}

@@

Observational studies and population censuses have become a cornerstone
of modern ecological research and conservation biology (Magurran et al.
2010; Hughes et al. 2017). Long-term data are necessary to both
understand population dynamics and to assess extinction risk of species.
Even though many time series may now be considered ``long-term'', most
are still short. Time series are typically short for a variety of
reasons. They are often coupled with an experiment, which may only last
a couple of years. In addition, short funding cycles make it difficult
to examine populations over longer periods of time (Hughes et al. 2017).
There are, of course, notable exceptions where long-term monitoring has
occurred, for example the continuous plankton recorder (Giron-Nava et
al. 2017).

Naturally, one might ask, how long of a time series is actually
necessary? This question has important implications for both research
and management. Scientifically, we need to know the time series length
required to address a specific question. Too short of a time series may
lead to wrong conclusions given large natural year-to-year variability
(citations). From a management perspective, it is important to
understand when population trend over time is actually meaningful or
not. The IUCN Red List Categories and Criteria suggest, under Criterion
A2, a species qualifies as vulnerable if it has experienced a 30\%
decline over 10 years, or 3 generations (IUCN 2012). For both scientific
and management questions, because sampling is typically expensive, we
also do not want to sample for longer than is necessary (Gerber,
DeMaster, and Kareiva 1999).

In determining the number of samples required for a particular
experiment four quantities are intricately linked: statistical power
(1-\(\beta\)), effect size, \(\alpha\), and sample size (Legg and Nagy
2006). The exact relationship between these quantities depends on the
specific statistical test. Formally, statistical power (\(1-\beta\)) is
one minus the probability of a type II error (\(\beta\)), or false
negative. In other words, it is a measure of how well you could detect a
trend given it is actually significant. Prior to an experiment, one
could set appropriate levels of power, significance threshold
(\(\alpha\)), and the effect size to estimate the sample size required
for the experiment. This approach, however, is not straight-forward for
a time series, or more complicated scenarios (P. C. Johnson et al.
2015), as data points are clearly non-independent.

For time series data, two general approaches can be used. First, as can
be done with experimental data, we can relax the definition of
statistical power (Bolker 2008). Instead of using classic statistical
tests (e.g.~t-test), and corresponding formulas for determining sample
size, simulation approaches can be used (Gerrodette 1987; Bolker 2008;
P. C. Johnson et al. 2015). Simple models can be simulated with
parameter values corresponding to a population of interest. Out of a
large number of simulations, power can be calculated as the fraction of
simulations that meet some criteria. The specific criteria depend on the
question at hand. Then, time series simulations of different lengths are
run to determine the minimum time series length required to be confident
that you will be able to answer your question (Bolker 2008). In addition
to using simulated data, empirical time series can also be used. We
usually do not have multiple replicates of similar populations, but we
can subsample an empirical time series (Gerber, DeMaster, and Kareiva
1999). Subsamples of different lengths can then be evaluated to see
which fraction of subsamples meet some criteria, again a measure of
statistical power. Similar to the simulation approach, we can use these
measures of power to determine the minimum time series required for a
particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized a model specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organism, 15 years was adequate to assess long-term trends
in abundance. However, these studies, and other past work, have
typically focused either on theoretical aspects of monitoring design or
focused on only a few species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
several questions. In the main manuscript, we estimate the minimum time
series length required to assess long-term changes in abundance via
simple linear regression. We then examine 878 population time series to
estimate this minimum time series length required for detecting
population trends. We expand on the results in the supplmentary material
by examining other questions of interest.

>>>>

To be submitted to: \emph{Biological Conservation} \vspace{3 mm}

@@

Long-term census data are necessary to better understand ecological
processes and make management decisions. However, census data are often
expensive, requiring a lot of time and resources. Little work has
addressed the question: when is a time series of census data long enough
to address a question of interest? We determine the minimum time series
length required to estimate significant increases or decreases in
population abundance. Importantly, we examine the ability to detect a
trend with a set level of statistical power, which is often neglected in
ecological time series analyses. Using a simulation approach, we
demonstrate how the minimum time series length required increases with
weaker trends in abundance and with higher population variability. In
addition, we examine 878 populations of vertebrate species to determine
the minimum time required to detect changes in their abundance. We show
that 10-15 years of continuous monitoring are required in order to
achieve a high level of statistical power. Similar to
simulation-approaches, the minimum time required for field census data
strongly depends on trend strength, population variability, and temporal
autocorrelation. These results point to the importance of sampling
populations over long periods of time. Time series less than 10-15 years
are likely under-powered and potentially misleading.

@@

Observational studies and population time series have become a
cornerstone of modern ecological research and conservation biology
(Magurran et al. 2010; Hughes et al. 2017). Long-term data are necessary
to both understand population dynamics and to assess species extinction
risk (citations). Even though many time series may now be considered
``long-term'' (e.g.~continuous plankton recorder Giron-Nava et al.
(2017)), most are still short. Time series are typically short for a
variety of reasons. They are often coupled with an experiment, which may
only last a couple of years. In addition, short funding cycles make it
difficult to examine populations over longer periods of time (Hughes et
al. 2017).

How long of a time series is actually necessary? This question has
important implications for both research and management. Scientists need
to know the time series length required to address a specific question.
A short time series may lead to wrong conclusions given large natural
year-to-year variability (citations). Managers need to know when action
is needed for a population. Therefore, they must understand when
population trend over time is actually meaningful. The IUCN Red List
Categories and Criteria suggest, under Criterion A2, a species qualifies
as vulnerable if it has experienced a 30\% decline over 10 years, or 3
generations (IUCN 2012). For both scientific and management questions,
because sampling is typically expensive, we also do not want to sample
for longer than is necessary (Gerber, DeMaster, and Kareiva 1999).

An important step in experimental design is to determine the number of
samples required. For any particular experiment four quantities are
intricately linked: significance level (\(\alpha\)), statistical power,
effect size, and sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. A type I error is a false positive, or incorrect
rejection of a true null hypothesis. For example, if a time series was
assessed as significantly increasing or decreasing---when there was no
true significant trend---this would be a false positive. The false
positive rate, or significance level (\(\alpha\)) is often set at 0.05.
A type II error (\(\beta\)) is failure to detect a true trend, or
failure to reject a false null hypothesis. Formally, statistical power
(\(1-\beta\)) is one minus the probability of a type II error
(\(\beta\)). The effect size is a measure of the difference between two
groups. Prior to an experiment, one could set appropriate levels of
power, significance level, and the effect size to estimate the sample
size required for the experiment. This approach, however, is not
straight-forward for a time series, or more complicated scenarios (P. C.
Johnson et al. 2015), as data are clearly non-independent.

For time series data, two general approaches to estimating sample size
are appropriate. Simulations can be designed for a specific population
and question (Bolker 2008; P. C. Johnson et al. 2015). Simple models can
be simulated with parameter values corresponding to a population of
interest (Gerrodette 1987). Statistical power can then be calculated as
the fraction of simulations that meet some criteria. The specific
criteria depend on the question at hand. For example, given a time
series, when is the slope from linear regression significantly different
from zero? In other words, when is the time series significantly
increasing or decreasing? It is then possible to determine how power
changes with a variable of interest. For example, time series are
simulated for different lengths of time. From these simulations, the
minimum time series length required is calculated to meet certain levels
of statistical significance and power (Bolker 2008).

In addition to using simulations, empirical time series can also be
used. Multiple replicates of similar populations are usually not
available, but it is possible to subsample an empirical time series
(Gerber, DeMaster, and Kareiva 1999). Subsamples of different lengths
can then be evaluated to see which fraction of subsamples meet some
criteria, again a measure of statistical power. Similar to the
simulation approach, we can use these measures of power to determine the
minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized a model specifically for a
threatened snail, \emph{Tasmaphena lamproides}. They found that for this
short-lived organism, 15 years was adequate to assess long-term trends
in abundance. In addition, Hatch (2003) examined seabird populations and
determined the minimum time required to achieve high statistical power
ranged between 11 and 69 years. However, these studies, and other past
work, have typically focused either on theoretical aspects of monitoring
design or focused on only a few species.

We use both simulations and a database of empirical time series to
determine the minimum number of years required to address several
questions. We estimate the minimum time series length required
(\(T_{min}\)) to assess long-term changes in abundance via simple linear
regression. First, we estimate \(T_{min}\) using a simulation approach.
Then we examine 878 population time series to estimate \(T_{min}\). In
the supplementary material, we determine \(T_{min}\) for related ideas:
calculating long-term growth rates, using more complicated population
models, varying statistical level and power, and the use of generalized
additive models.
